---
title: "Architecture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(herald)

extdata <- function(file) {
  candidates <- c(
    system.file("extdata", file, package = "herald"),
    file.path("..", "inst", "extdata", file),
    file.path("inst", "extdata", file)
  )
  hits <- candidates[nzchar(candidates) & file.exists(candidates)]
  if (!length(hits)) stop("Cannot find extdata file: ", file)
  hits[[1L]]
}
```

`herald` is organized as a set of small layers. Each layer can be used
directly, but the usual production path flows from data ingest to validation
reports.

```
6  reports              report(), write_report_html(), write_report_xlsx()
5  validation engine    validate(), rule_catalog(), supported_standards()
4  rule operations      op_* functions registered in R/ops-*.R
3  dictionaries         ct_provider(), srs_provider(), custom_provider()
2  specifications       as_herald_spec(), herald_spec(), read_define_xml()
1  metadata stamping    apply_spec()
0  dataset I/O          read_xpt(), read_json(), read_parquet()
```

## Dataset contract

Datasets are normal R data frames. Metadata is carried in attributes:

```{r attrs}
dm <- readRDS(extdata("dm.rds"))
spec <- readRDS(extdata("sdtm-spec.rds"))
dm <- apply_spec(list(DM = dm), spec)$DM

attr(dm, "label")
attr(dm$USUBJID, "label")
attr(dm$USUBJID, "sas.length")
attr(dm$USUBJID, "xpt_type")
```

This keeps the package compatible with base R, `dplyr`, `data.table`, and
other data-frame tooling while still preserving submission metadata for XPT
and report writers.

## Specification objects

`as_herald_spec()` is the compact constructor for dataset and variable
metadata. `herald_spec()` is the richer constructor used by Define-XML.

```{r spec}
simple_spec <- as_herald_spec(
  ds_spec = data.frame(
    dataset = "DM",
    label = "Demographics",
    class = "SPECIAL PURPOSE",
    stringsAsFactors = FALSE
  ),
  var_spec = data.frame(
    dataset = "DM",
    variable = c("STUDYID", "USUBJID"),
    label = c("Study Identifier", "Unique Subject Identifier"),
    type = c("text", "text"),
    stringsAsFactors = FALSE
  )
)

simple_spec
```

## Rule catalog

Compiled rules live under `inst/rules`. `rule_catalog()` flattens the compiled
corpus for browsing, filtering, and coverage checks.

```{r catalog}
catalog <- rule_catalog()

nrow(catalog)
head(catalog[, c("rule_id", "standard", "authority", "severity", "has_predicate")])
```

`supported_standards()` summarizes predicate coverage by standard and authority.

```{r supported}
supported_standards()
```

## Operation registry

Rule predicates are implemented by operation functions in `R/ops-*.R`.
Operations return logical vectors with this convention:

| Return value | Meaning |
|---|---|
| `TRUE` | Rule fires for that row or dataset |
| `FALSE` | Rule passes |
| `NA` | Advisory or skipped because required context is unavailable |

Every operation is registered with metadata so the rule engine can dispatch by
name and report unresolved references.

## Error contract

Package errors inherit from `herald_error`. Specific subclasses distinguish
input, file, spec, rule, validation, Define-XML, report, and runtime failures.
This lets pipelines catch broadly while tests pin precise classes.

```{r error-contract}
tryCatch(
  validate(quiet = TRUE),
  herald_error = function(e) class(e)
)
```

## Extension points

The safest extension points are:

| Need | Extension point |
|---|---|
| New controlled terminology source | `new_dict_provider()` or `custom_provider()` |
| Sponsor-specific dictionary | `register_dictionary()` or `validate(dictionaries = ...)` |
| Severity policy | `validate(severity_map = ...)` |
| Output automation | `report()` or specific report writers |
| Define-XML authoring | `herald_spec()` plus `write_define_xml()` |

Internal rule operation authoring is more powerful, but it changes validation
semantics and should be paired with focused fixtures.

## Development checks

The package-level quality gate is:

```r
devtools::document()
devtools::test()
devtools::check(args = "--no-manual")
```

For rule development, also track fixture coverage with:

```r
source("tools/fixture-coverage.R")
```

See `vignette("rule-coverage")` for the current fixture coverage report.
