---
title: "Best Practices for Validation Projects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Best Practices for Validation Projects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(herald)

extdata <- function(file) {
  candidates <- c(
    system.file("extdata", file, package = "herald"),
    file.path("..", "inst", "extdata", file),
    file.path("inst", "extdata", file)
  )
  hits <- candidates[nzchar(candidates) & file.exists(candidates)]
  if (!length(hits)) stop("Cannot find extdata file: ", file)
  hits[[1L]]
}
```

This guide shows how to organize `herald` in a production validation project:
metadata first, deterministic inputs, explicit rule selection, reproducible
dictionaries, and reports that can be reviewed by programmers and auditors.

## Recommended project layout

Keep submitted data, metadata, validation outputs, and project setup separate.

```text
study/
  data/
    sdtm/
      dm.xpt
      ae.xpt
    adam/
      adsl.xpt
      adae.xpt
  metadata/
    define.xml
    spec.rds
  dictionaries/
    meddra/
    whodrug/
  programs/
    validate-sdtm.R
    validate-adam.R
  qc/
    sdtm-report.html
    sdtm-report.xlsx
    sdtm-report.json
```

The validation program should read only from `data/`, `metadata/`, and
`dictionaries/`, then write only to `qc/`. That makes reruns predictable.

## Metadata first

Stamp metadata before writing transport files and before validation. This
keeps labels, lengths, and expected types synchronized across XPT,
Dataset-JSON, Define-XML, and reports.

```{r metadata-first}
dm <- readRDS(extdata("dm.rds"))
spec <- readRDS(extdata("sdtm-spec.rds"))

dm <- apply_spec(dm, spec)

attr(dm, "label")
attr(dm$USUBJID, "label")
attr(dm$USUBJID, "sas.length")
```

## Validate submitted artifacts

In production, validate the files you plan to submit, not an earlier in-memory
object. A good pattern is:

1. Stamp metadata.
2. Write XPT or Dataset-JSON.
3. Read the written files back.
4. Validate the reread files.
5. Write all report formats.

```{r validate-artifacts}
root <- file.path(tempdir(), "herald-best-practices")
data_dir <- file.path(root, "data", "sdtm")
qc_dir <- file.path(root, "qc")
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(qc_dir, recursive = TRUE, showWarnings = FALSE)

write_xpt(dm, file.path(data_dir, "dm.xpt"))

result <- validate(
  path  = data_dir,
  rules = character(0),
  quiet = TRUE
)

result$datasets_checked
```

## Make rule selection explicit

Do not let production programs hide the rule policy. Define rule filters near
the top of the program.

```{r rule-policy}
catalog <- rule_catalog()

rule_policy <- list(
  standards   = c("SDTM-IG", "Define-XML"),
  authorities = c("CDISC", "FDA"),
  severity_map = c("Medium" = "High")
)

head(catalog[catalog$standard %in% rule_policy$standards,
             c("rule_id", "standard", "authority", "severity")])
```

Use the same object in the validation call:

```{r policy-run}
policy_result <- validate(
  path = data_dir,
  standards = rule_policy$standards,
  authorities = rule_policy$authorities,
  severity_map = rule_policy$severity_map,
  quiet = TRUE
)

policy_result$rules_total
```

## Keep dictionaries reproducible

Pass dictionary providers directly to `validate()` when you want the run to be
self-contained. Use session registration for interactive work.

```{r dictionaries}
dictionaries <- list(
  "ct-sdtm" = ct_provider("sdtm")
)

dict_result <- validate(
  path = data_dir,
  rules = character(0),
  dictionaries = dictionaries,
  quiet = TRUE
)

dict_result$skipped_refs
```

For licensed dictionaries such as MedDRA or WHODrug, store the source version
in the project metadata and construct providers in one setup block.

```r
dictionaries <- list(
  "ct-sdtm" = ct_provider("sdtm", version = "bundled"),
  "meddra"  = meddra_provider("dictionaries/meddra", version = "27.0"),
  "whodrug" = whodrug_provider("dictionaries/whodrug", version = "2024-03")
)
```

## Report every run

Write all three report formats from the same `herald_result`.

```{r reports}
html <- file.path(qc_dir, "sdtm-report.html")
xlsx <- file.path(qc_dir, "sdtm-report.xlsx")
json <- file.path(qc_dir, "sdtm-report.json")

report(result, html)
report(result, xlsx)
report(result, json)

file.exists(c(html, xlsx, json))
```

Use each format for a different audience:

| Format | Best use |
|---|---|
| HTML | Programmer and study-team review |
| XLSX | Triage, filtering, comments, issue assignment |
| JSON | CI gates, dashboards, audit archives |

## Batch validation

A study usually has multiple packages: SDTM, ADaM, and sometimes standalone
Define-XML checks. Use a small metadata table and loop over it.

```{r batch, eval = FALSE}
jobs <- data.frame(
  package = c("sdtm", "adam"),
  path    = c("data/sdtm", "data/adam"),
  html    = c("qc/sdtm-report.html", "qc/adam-report.html"),
  xlsx    = c("qc/sdtm-report.xlsx", "qc/adam-report.xlsx"),
  json    = c("qc/sdtm-report.json", "qc/adam-report.json"),
  stringsAsFactors = FALSE
)

for (i in seq_len(nrow(jobs))) {
  job <- jobs[i, ]
  res <- validate(path = job$path, quiet = TRUE)
  report(res, job$html)
  report(res, job$xlsx)
  report(res, job$json)
}
```

## CI gate

A simple CI policy can fail on high-impact fired findings while still saving
the full report set for review.

```{r ci-gate}
is_blocking <- function(result) {
  f <- result$findings
  if (!nrow(f)) return(FALSE)
  any(
    f$status == "fired" &
      f$severity %in% c("Reject", "Error", "High"),
    na.rm = TRUE
  )
}

is_blocking(result)
```

In CI, write reports first, then stop:

```r
report(result, "qc/report.html")
report(result, "qc/report.xlsx")
report(result, "qc/report.json")

if (is_blocking(result)) {
  stop("Blocking validation findings found. See qc/report.html.")
}
```

## QC checklist

Before a validation run is considered review-ready:

| Check | Command |
|---|---|
| Package loads from a clean R session | `library(herald)` |
| Rule corpus is inspectable | `supported_standards()` |
| Datasets can be reread from disk | `validate(path = "data/sdtm")` |
| Dictionary versions are explicit | `ct_info("sdtm")`, provider `$info()` |
| Reports are written from one result | `report(result, ...)` |
| Skipped references are reviewed | `result$skipped_refs` |
| Blocking findings are gated | `is_blocking(result)` |

The key habit is to keep the validation result as the source of truth. Reports,
CI gates, dashboards, and manual issue triage should all derive from the same
`herald_result` object.
