---
title: "Dataset I/O and Format Conversion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dataset I/O and Format Conversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(herald)

extdata <- function(file) {
  candidates <- c(
    system.file("extdata", file, package = "herald"),
    file.path("..", "inst", "extdata", file),
    file.path("inst", "extdata", file)
  )
  hits <- candidates[nzchar(candidates) & file.exists(candidates)]
  if (!length(hits)) stop("Cannot find extdata file: ", file)
  hits[[1L]]
}
```

`herald` treats clinical data frames as ordinary R data frames plus a small
attribute contract:

- dataset label: `attr(x, "label")`
- dataset name: `attr(x, "dataset_name")`
- variable label: `attr(x$VAR, "label")`
- XPT storage length: `attr(x$VAR, "sas.length")`
- XPT type: `attr(x$VAR, "xpt_type")`

Those attributes are used by the XPT, Dataset-JSON, Parquet, and reporting
layers.

## Prepare a dataset

```{r prepare}
dm <- readRDS(extdata("dm.rds"))
spec <- readRDS(extdata("sdtm-spec.rds"))
dm <- apply_spec(dm, spec)

attr(dm, "label")
attr(dm$USUBJID, "label")
```

## XPT

`write_xpt()` writes SAS transport files in pure R. `read_xpt()` reads them
back and restores dataset and column metadata.

```{r xpt}
io_dir <- file.path(tempdir(), "herald-io")
dir.create(io_dir, showWarnings = FALSE)

xpt <- file.path(io_dir, "dm.xpt")
write_xpt(dm, xpt)

dm_xpt <- read_xpt(xpt)

nrow(dm_xpt)
attr(dm_xpt, "label")
attr(dm_xpt$USUBJID, "label")
```

## Dataset-JSON

Dataset-JSON is useful for APIs, review automation, and text-friendly
interchange.

```{r json}
json <- file.path(io_dir, "dm.json")
write_json(dm, json)

dm_json <- read_json(json)

identical(nrow(dm), nrow(dm_json))
attr(dm_json$USUBJID, "label")
```

## Parquet

Parquet support uses the optional `arrow` package. Keep the code guarded in
portable vignettes and CI jobs.

```{r parquet, eval = requireNamespace("arrow", quietly = TRUE)}
parquet <- file.path(io_dir, "dm.parquet")
write_parquet(dm, parquet)

dm_parquet <- read_parquet(parquet)
nrow(dm_parquet)
```

## Conversion

`convert_dataset()` infers formats from file extensions. Use `from =` and
`to =` when extensions are nonstandard.

```{r conversion}
xpt2  <- file.path(io_dir, "dm-roundtrip.xpt")
json2 <- file.path(io_dir, "dm-roundtrip.json")

convert_dataset(xpt, json2)
convert_dataset(json2, xpt2)

file.exists(c(json2, xpt2))
```

Same-format conversion is also valid. It is useful when you want to normalize
metadata or smoke-test read/write symmetry.

```{r same-format}
json3 <- file.path(io_dir, "dm-copy.json")
convert_dataset(json2, json3)
file.exists(json3)
```

## Validation after I/O

The validation layer is format-blind after data is loaded.

```{r validate-after-io}
dm_loaded <- read_xpt(xpt)

result <- validate(
  files = list(DM = dm_loaded),
  rules = character(0),
  quiet = TRUE
)

result$datasets_checked
```

The recommended production pattern is to stamp metadata once with
`apply_spec()`, write transport files, read them back in CI, and validate the
files that will actually be submitted.
