---
title: "Define-XML 2.1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Define-XML 2.1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(herald)

extdata <- function(file) {
  candidates <- c(
    system.file("extdata", file, package = "herald"),
    file.path("..", "inst", "extdata", file),
    file.path("inst", "extdata", file)
  )
  hits <- candidates[nzchar(candidates) & file.exists(candidates)]
  if (!length(hits)) stop("Cannot find extdata file: ", file)
  hits[[1L]]
}
```

`herald` has a first-class Define-XML path:

```
herald_spec -> write_define_xml() -> read_define_xml() -> validate()
```

The writer supports Define-XML 2.1 structures such as study metadata,
datasets, variables, codelists, methods, comments, documents, and ADaM ARM
tables.

## Build a rich spec

```{r build-spec}
spec <- herald_spec(
  study = data.frame(
    attribute = c("StudyName", "StudyDescription", "ProtocolName"),
    value     = c("PILOT01", "Pilot study", "PILOT01"),
    stringsAsFactors = FALSE
  ),
  ds_spec = data.frame(
    dataset = c("DM", "AE"),
    label   = c("Demographics", "Adverse Events"),
    class   = c("SPECIAL PURPOSE", "EVENTS"),
    stringsAsFactors = FALSE
  ),
  var_spec = data.frame(
    dataset   = c("DM", "DM", "DM", "AE", "AE", "AE"),
    variable  = c("STUDYID", "USUBJID", "SEX", "STUDYID", "USUBJID", "AETERM"),
    label     = c("Study Identifier", "Unique Subject Identifier", "Sex",
                  "Study Identifier", "Unique Subject Identifier", "Reported Term"),
    data_type = c("text", "text", "text", "text", "text", "text"),
    length    = c("12", "40", "1", "12", "40", "200"),
    stringsAsFactors = FALSE
  ),
  codelist = data.frame(
    codelist_id   = c("CL.SEX", "CL.SEX"),
    name          = c("Sex", "Sex"),
    data_type     = c("text", "text"),
    term          = c("M", "F"),
    decoded_value = c("Male", "Female"),
    stringsAsFactors = FALSE
  ),
  methods = data.frame(
    method_id   = "MT.AGE",
    name        = "Derive AGE",
    type        = "Computation",
    description = "AGE is derived from reference start date and birth date.",
    stringsAsFactors = FALSE
  )
)

spec
```

## Write XML

Use `validate = FALSE` while drafting incomplete metadata. Turn validation on
when the spec is ready for release.

```{r write-xml, eval = requireNamespace("xml2", quietly = TRUE)}
define_dir <- file.path(tempdir(), "herald-define")
dir.create(define_dir, showWarnings = FALSE)

xml_path <- file.path(define_dir, "define.xml")
write_define_xml(spec, xml_path, validate = FALSE)

file.exists(xml_path)
readLines(xml_path, n = 5)
```

## Read XML back

`read_define_xml()` returns a `herald_define` object. The object exposes the
same data-frame slots used by validation and Define-XML dependent rules.

```{r read-xml, eval = requireNamespace("xml2", quietly = TRUE)}
define <- read_define_xml(xml_path)

define
define$ds_spec[, c("dataset", "label")]
define$var_spec[, c("dataset", "variable", "label")]
```

## Render reviewer HTML

`write_define_html()` renders a reviewer-friendly HTML page from the spec.

```{r write-html, eval = requireNamespace("xml2", quietly = TRUE)}
html_path <- file.path(define_dir, "define.html")
write_define_html(spec, html_path, define_xml = xml_path)

file.exists(html_path)
```

## Validate with Define-XML context

Pass the parsed Define-XML object to `validate()` when rule execution needs
metadata that is not present in datasets alone.

```{r validate-define, eval = requireNamespace("xml2", quietly = TRUE)}
dm <- readRDS(extdata("dm.rds"))
dm <- apply_spec(dm, spec)

result <- validate(
  files  = list(dm, define = define),
  define = define,
  rules  = character(0),
  quiet  = TRUE
)

result$datasets_checked
```

## Practical guidance

Keep Define-XML authoring in the same pipeline as dataset writing:

1. Build or read the `herald_spec`.
2. Apply the spec to datasets.
3. Write XPT or Dataset-JSON.
4. Write Define-XML and HTML.
5. Read the written Define-XML back.
6. Validate the submitted files with the parsed Define-XML object.

This catches drift between the metadata source and the actual submission
artifacts before reviewers see it.
