---
title: "Controlled Terminology and Dictionary Providers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Controlled Terminology and Dictionary Providers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(herald)
```

Clinical validation depends on reference data: CDISC controlled terminology,
SRS, MedDRA, WHODrug, LOINC, SNOMED, and sponsor dictionaries. `herald`
uses one interface for all of them: a dictionary provider.

## Bundled CDISC CT

The package ships bundled SDTM and ADaM controlled terminology.

```{r ct-info}
ct_info("sdtm")
ct_info("adam")
```

Load the raw CT when you need direct inspection:

```{r load-ct}
sdtm_ct <- load_ct("sdtm")

length(sdtm_ct)
names(sdtm_ct)[1:10]

sdtm_ct[["NY"]]$terms[, c("submissionValue", "preferredTerm")]
```

## CT as a provider

Use `ct_provider()` when rule operations need dictionary lookup behavior.

```{r ct-provider}
p <- ct_provider("sdtm")

p
p$contains(c("Y", "N", "UNKNOWN"), field = "NY")
p$contains("y", field = "NY", ignore_case = TRUE)

p$lookup(c("Y", "N"), field = "NY")[, c("submissionValue", "preferredTerm")]
```

## Registering providers

Register providers globally for the R session, or pass them directly to
`validate(dictionaries = ...)` for one run.

```{r register}
register_dictionary("ct-sdtm", ct_provider("sdtm"))
list_dictionaries()
unregister_dictionary("ct-sdtm")
```

One-run registration avoids session state:

```{r one-run}
dm <- data.frame(USUBJID = "S1", stringsAsFactors = FALSE)

result <- validate(
  files = list(dm),
  rules = character(0),
  dictionaries = list("ct-sdtm" = ct_provider("sdtm")),
  quiet = TRUE
)

result
```

## Custom dictionaries

Use `custom_provider()` when your dictionary is already in a data frame.

```{r custom-provider}
sponsor_terms <- data.frame(
  VISIT = c("SCREENING", "BASELINE", "WEEK 1"),
  stringsAsFactors = FALSE
)

visit_provider <- custom_provider(
  sponsor_terms,
  name = "sponsor-visits",
  fields = "VISIT"
)

visit_provider
visit_provider$contains(c("BASELINE", "EARLY TERMINATION"), field = "VISIT")
```

For full control, build a provider directly with `new_dict_provider()`.

```{r new-provider}
allowed <- c("LOW", "MEDIUM", "HIGH")

risk_provider <- new_dict_provider(
  name      = "risk-levels",
  version   = "2026-04",
  source    = "sponsor",
  license   = "internal",
  size_rows = length(allowed),
  fields    = "RISK",
  contains  = function(value, field = NULL, ignore_case = FALSE) {
    value <- as.character(value)
    if (isTRUE(ignore_case)) {
      return(toupper(value) %in% toupper(allowed))
    }
    value %in% allowed
  }
)

risk_provider$contains(c("LOW", "UNKNOWN"), field = "RISK")
```

## External dictionaries

The package also exposes factories for common clinical dictionaries:

| Provider | Function |
|---|---|
| SRS | `download_srs()`, `srs_provider()` |
| MedDRA | `meddra_provider()` |
| WHODrug | `whodrug_provider()` |
| LOINC | `loinc_provider()` |
| SNOMED | `snomed_provider()` |

Those functions expect licensed or downloaded source files and are usually
configured in project-specific setup scripts rather than inside analysis code.
