---
title: "Getting Started with herald"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with herald}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(herald)

extdata <- function(file) {
  candidates <- c(
    system.file("extdata", file, package = "herald"),
    file.path("..", "inst", "extdata", file),
    file.path("inst", "extdata", file)
  )
  hits <- candidates[nzchar(candidates) & file.exists(candidates)]
  if (!length(hits)) stop("Cannot find extdata file: ", file)
  hits[[1L]]
}
```

`herald` is a clinical submission validation package. It is designed for
R-native pipelines that need to read submission datasets, attach CDISC
metadata, execute conformance rules, and produce auditable reports.

The current package has four core objects:

| Object | What it represents |
|---|---|
| `data.frame` with attributes | A clinical dataset with labels, lengths, and XPT type metadata |
| `herald_spec` | Dataset and variable metadata used before validation |
| `herald_define` | A Define-XML document parsed into R data frames |
| `herald_result` | Validation findings, rule snapshot, dataset metadata, and skipped references |

## The workflow

Most workflows follow the same shape:

```
read data -> build/read spec -> apply spec -> validate -> report
```

Use the pilot data bundled with the package:

```{r load-pilot}
dm <- readRDS(extdata("dm.rds"))
spec <- readRDS(extdata("sdtm-spec.rds"))

dim(dm)
names(dm)[1:6]
spec
```

## Apply metadata

`apply_spec()` stamps dataset-level and variable-level metadata. The result is
still a regular data frame, so existing R code keeps working.

```{r apply-spec}
dm <- apply_spec(list(DM = dm), spec)$DM

attr(dm, "label")
attr(dm$USUBJID, "label")
attr(dm$USUBJID, "sas.length")
attr(dm$USUBJID, "xpt_type")
```

## Validate

`validate()` accepts either a directory path or a named list of data frames.
For a first smoke test, use `rules = character(0)` to exercise the result
object and reporting path without running the full compiled corpus.

```{r validate-smoke}
result <- validate(
  files = list(DM = dm),
  rules = character(0),
  quiet = TRUE
)

result
names(result)
```

Run selected rules by passing rule identifiers. To inspect what is available,
use `rule_catalog()`.

```{r selected-rules}
catalog <- rule_catalog()
head(catalog[, c("rule_id", "standard", "authority", "severity", "has_predicate")])

selected <- head(catalog$rule_id[catalog$has_predicate], 3)
selected

selected_result <- validate(
  files = list(DM = dm),
  rules = selected,
  quiet = TRUE
)

selected_result$rules_total
selected_result$datasets_checked
```

## Write reports

Reports are generated from a `herald_result`. The generic `report()` dispatches
from the file extension.

```{r report-files}
out_dir <- file.path(tempdir(), "herald-getting-started")
dir.create(out_dir, showWarnings = FALSE)

html_path <- file.path(out_dir, "validation-report.html")
json_path <- file.path(out_dir, "validation-report.json")
xlsx_path <- file.path(out_dir, "validation-report.xlsx")

report(selected_result, html_path)
report(selected_result, json_path)
report(selected_result, xlsx_path)

file.exists(c(html_path, json_path, xlsx_path))
```

## Convert dataset formats

`convert_dataset()` converts among XPT, Dataset-JSON, and Parquet. Parquet
requires the optional `arrow` package.

```{r convert-xpt-json}
xpt_path  <- file.path(out_dir, "dm.xpt")
json_path <- file.path(out_dir, "dm.json")

write_xpt(dm, xpt_path)
convert_dataset(xpt_path, json_path)

dm_from_json <- read_json(json_path)
identical(nrow(dm), nrow(dm_from_json))
```

## Where to go next

- `vignette("validation-reporting")` for rule filters, severity overrides, and reports.
- `vignette("data-io")` for XPT, Dataset-JSON, Parquet, and conversions.
- `vignette("define-xml")` for Define-XML 2.1 read/write round-trips.
- `vignette("dictionaries")` for controlled terminology and external dictionaries.
- `vignette("architecture")` for package layers and extension points.
