---
title: "Validation and Reporting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation and Reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(herald)

extdata <- function(file) {
  candidates <- c(
    system.file("extdata", file, package = "herald"),
    file.path("..", "inst", "extdata", file),
    file.path("inst", "extdata", file)
  )
  hits <- candidates[nzchar(candidates) & file.exists(candidates)]
  if (!length(hits)) stop("Cannot find extdata file: ", file)
  hits[[1L]]
}
```

`validate()` is the main entry point. It returns a `herald_result`: a stable
object containing findings, the rule catalog snapshot, dataset metadata,
skipped references, timing, and the detected submission profile.

## Inputs

Validation can start from in-memory data frames:

```{r memory-input}
dm <- readRDS(extdata("dm.rds"))
spec <- readRDS(extdata("sdtm-spec.rds"))

dm <- apply_spec(list(DM = dm), spec)$DM

r0 <- validate(
  files = list(DM = dm),
  rules = character(0),
  quiet = TRUE
)

r0
```

Or from a directory of files:

```{r path-input}
submission_dir <- file.path(tempdir(), "herald-validation")
dir.create(submission_dir, showWarnings = FALSE)

write_xpt(dm, file.path(submission_dir, "dm.xpt"))

r1 <- validate(
  path  = submission_dir,
  rules = character(0),
  quiet = TRUE
)

r1$datasets_checked
```

## Rule selection

Use `rule_catalog()` to find rule IDs, then pass those IDs to `validate()`.
`authorities` and `standards` give broader filters.

```{r rule-selection}
catalog <- rule_catalog()

table(catalog$standard, catalog$authority)
head(catalog[catalog$has_predicate, c("rule_id", "standard", "severity", "message")])

rule_ids <- head(catalog$rule_id[catalog$has_predicate], 5)

r2 <- validate(
  files = list(DM = dm),
  rules = rule_ids,
  quiet = TRUE
)

r2$rules_total
r2$rules_applied
```

For exploratory runs, `supported_standards()` gives a corpus-level summary.

```{r standards}
supported_standards()
```

## Findings

Findings are stored in a tibble. The most useful columns for triage are
`rule_id`, `severity`, `status`, `dataset`, `variable`, `row`, and `message`.

```{r findings}
names(r2$findings)

if (nrow(r2$findings)) {
  r2$findings[, c("rule_id", "severity", "status", "dataset", "variable", "row", "message")]
} else {
  "No findings emitted by the selected rules."
}
```

Filter with base R or your preferred data-frame toolkit:

```{r filter-findings}
fired <- r2$findings[r2$findings$status == "fired", , drop = FALSE]
high  <- r2$findings[r2$findings$severity %in% c("Reject", "High", "Error"), , drop = FALSE]

nrow(fired)
nrow(high)
```

## Severity overrides

Sponsors often need study-specific severity policies. `severity_map` can
override by exact rule ID, by regular expression, or by severity category.

```{r severity-map}
r3 <- validate(
  files = list(DM = dm),
  rules = rule_ids,
  severity_map = c("Medium" = "High"),
  quiet = TRUE
)

if (nrow(r3$findings)) {
  unique(r3$findings[, c("rule_id", "severity", "severity_override")])
} else {
  "No emitted findings to override in this small run."
}
```

## Reports

Use `report()` when you want extension-based dispatch. Use the specific
writers when you want explicit output.

```{r reports}
report_dir <- file.path(tempdir(), "herald-reports")
dir.create(report_dir, showWarnings = FALSE)

html <- file.path(report_dir, "report.html")
xlsx <- file.path(report_dir, "report.xlsx")
json <- file.path(report_dir, "report.json")

write_report_html(r2, html)
write_report_xlsx(r2, xlsx)
write_report_json(r2, json)

file.info(c(html, xlsx, json))[, c("size", "mtime")]
```

## Skipped references

Some rules require datasets or dictionaries that are absent. Those are reported
as skipped references rather than hidden as silent passes.

```{r skipped-refs}
str(r2$skipped_refs, max.level = 2)
```

The result object keeps enough context for both automated pipelines and human
review. A CI job can fail on `nrow(result$findings) > 0`, while a clinical
programmer can open the HTML or XLSX report for triage.
