---
title: "Frequently Asked Questions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Frequently Asked Questions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(herald)
```

Short answers to the questions new users ask most often. Each entry
shows the smallest runnable code that resolves the issue.

## Why does my `validate()` return 0 rules?

Three common causes:

1. **Empty rule filter.** `rules = character(0)` is a deliberate
   smoke-test mode that exercises the result object without running
   anything. Drop the argument to run the full corpus.

2. **Filter excludes every rule.** Combining `standards =`,
   `authorities =`, and `rules =` ANDs the conditions. Inspect with
   `rule_catalog()` first.

3. **Datasets do not match any rule scope.** Some rules are
   dataset-scoped (e.g. AE-only rules). If you only pass DM, AE-only
   rules are silently out of scope.

```{r zero-rules}
# Smoke-test mode: rules_total = 0 by design
r <- validate(files = dm, rules = character(0), quiet = TRUE)
r$rules_total

# Run the full corpus
r <- validate(files = list(DM = apply_spec(dm, sdtm_spec)), quiet = TRUE)
r$rules_total
```

## How do I run only SDTM rules?

Use `standards = "SDTM-IG"`. Any combination of standards is allowed.

```{r sdtm-only}
r <- validate(
  files = list(DM = apply_spec(dm, sdtm_spec)),
  standards = "SDTM-IG",
  quiet = TRUE
)
r$rules_total
```

The available standard names are:

```{r supported}
supported_standards()$standard
```

## How do I add MedDRA?

Build a provider with `meddra_provider()`, register it, then run.
The provider expects a directory of MedDRA ASCII files (`mdhier.asc`,
`llt.asc`, etc.) and a version string for reproducibility.

```{r meddra, eval = FALSE}
meddra <- meddra_provider("dictionaries/meddra", version = "27.0")
register_dictionary("meddra", meddra)

result <- validate(path = "data/adam", quiet = TRUE)
```

If MedDRA is not registered when a rule needs it, the rule is
skipped and `result$skipped_refs$dictionaries$meddra` will explain
how to wire it up.

## Why is my Define-XML rejected?

`read_define_xml()` validates against the Define-XML 2.1 schema and
the herald conformance ruleset. Common rejections:

- **Missing namespace declaration** on the root element.
- **OID collisions** -- e.g. two `MetaDataVersion/ItemDef` items
  sharing the same `OID`.
- **Codelist references that point to nothing** -- a variable cites
  `CL.SEX` but no `<CodeList OID="CL.SEX">` block exists.

Run `validate_spec()` directly on the parsed object to see every
finding before it derails downstream code:

```{r define-validate, eval = FALSE}
d <- read_define_xml("metadata/define.xml")
findings <- validate_spec(d)
findings[findings$severity == "High", ]
```

## Why are my AE / VS rules being skipped?

Cross-dataset rules need their reference datasets present in the
same `validate()` call. If you pass only `AE`, any rule that needs
`DM` is recorded in `result$skipped_refs$datasets`.

```{r skipped-refs}
r <- validate(
  files = list(AE = adae),
  standards = "SDTM-IG",
  quiet = TRUE
)
str(r$skipped_refs$datasets, max.level = 1)
```

Provide every related domain together:

```{r related-domains}
r <- validate(
  files = list(
    DM = apply_spec(dm,  sdtm_spec),
    AE = adae
  ),
  standards = "SDTM-IG",
  quiet = TRUE
)
r$datasets_checked
```

## How do I keep severities under control across runs?

Use `severity_map=` to rewrite labels at run time, and store the
mapping in your project config (not in the rule catalog).

```{r severity-policy}
sponsor_policy <- c("Medium" = "High")

r <- validate(
  files = list(DM = apply_spec(dm, sdtm_spec)),
  standards = "SDTM-IG",
  severity_map = sponsor_policy,
  quiet = TRUE
)
table(r$findings$severity)
```

## Can I read SAS XPT without `haven` / SAS?

Yes. `read_xpt()` is pure R and supports XPT v5. It preserves
labels, lengths, and SAS types.

```{r xpt-pure-r}
out <- file.path(tempdir(), "herald-faq")
dir.create(out, showWarnings = FALSE)

write_xpt(apply_spec(dm, sdtm_spec), file.path(out, "dm.xpt"))
dm2 <- read_xpt(file.path(out, "dm.xpt"))

attr(dm2, "label")
attr(dm2$USUBJID, "label")
attr(dm2$USUBJID, "sas.length")
```

## Can I write a Define-XML from scratch?

Build a `herald_spec` (use `herald_spec()` for full control) and
write it. Round-tripping a parsed Define-XML through
`write_define_xml()` is also supported and preserves codelists,
methods, comments, and ARM metadata.

```{r write-define, eval = FALSE}
spec <- herald_spec(
  ds_spec  = ds_spec_df,
  var_spec = var_spec_df
)
write_define_xml(spec, "metadata/define.xml")
```

## How do I do XPT -> Dataset-JSON -> Parquet conversion?

`convert_dataset()` infers source and target formats from the file
extensions. Parquet support requires the optional `arrow` package.

```{r convert}
out <- file.path(tempdir(), "herald-faq-convert")
dir.create(out, showWarnings = FALSE)

write_xpt(apply_spec(dm, sdtm_spec), file.path(out, "dm.xpt"))
convert_dataset(file.path(out, "dm.xpt"), file.path(out, "dm.json"))
file.exists(file.path(out, "dm.json"))
```

## How do I tell which rule fired on which row?

Inspect the `findings` table on the result. Each row of `findings`
points at the originating rule and dataset.

```{r findings, eval = FALSE}
r <- validate(path = "data/sdtm", quiet = TRUE)
head(r$findings[, c("rule_id", "dataset", "severity", "status")])
```

## How do I run herald in CI?

Any CI that runs `Rscript` can run herald. A minimal GitHub Actions
step looks like:

```yaml
- uses: r-lib/actions/setup-r@v2
- run: Rscript -e 'install.packages("pak"); pak::pak("vthanik/herald")'
- run: Rscript -e 'herald::validate(path="data/sdtm") |>
                   herald::report("qc/report.html")'
```

Use the `is_blocking()` helper from the *Cookbook* article to fail
the job on high-impact fired findings while still uploading the
reports as artifacts.

## How does herald differ from Pinnacle 21?

See the dedicated `migrating-from-p21` article for a full mapping.
The short version:

- Same rule sources (SDTM-IG, ADaM-IG, Define-XML 2.1, CORE).
- Pure R, no JVM. Installs from CRAN-style sources.
- Reports in HTML, XLSX, and JSON from the same result object.
- Dictionary and rule configuration live in code, not in a GUI.

## Where do I report bugs?

Open an issue at <https://github.com/vthanik/herald/issues> with a
minimal reproducible example. The pilot datasets bundled with the
package (`dm`, `adsl`, `adae`, `advs`) are a good starting point.

## Where to go next

- `cookbook` -- ready-to-run recipes against the bundled pilot data.
- `extending-herald` -- write your own dictionary or operator.
- `migrating-from-p21` -- side-by-side mapping from Pinnacle 21.
