---
title: "Operator catalog"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Operator catalog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)
library(herald)
```

An **operator** (or "op") is the executable kernel of a herald rule. Each
op is a function `op_<name>(data, ctx, ...)` that returns a logical vector
the same length as `nrow(data)`: `TRUE` means the record fires the rule
(a finding is emitted), `FALSE` means the record passes, and `NA` means
the check is indeterminate (recorded as advisory). Rules in the YAML
corpus refer to ops by name; the validation engine looks them up in the
operator registry at run time.

The catalog below is generated at build time from the live registry.
For the broader picture of how ops fit into the layer stack, see the
[Architecture](../architecture.html) article. For guidance on writing
new ops, see the operator pattern notes in `.claude/rules/r-code.md`.

Ops are grouped by `kind`:

- **set** -- membership / codelist / domain checks.
- **compare** -- value comparisons against a literal or expression.
- **existence** -- presence, absence, completeness, uniqueness.
- **temporal** -- date / datetime ordering and ISO 8601 conformance.
- **cross** -- references another dataset via `.ref_ds(ctx, name)`.
- **string** -- regex, length, casing, whitespace.

```{r build-catalog}
ops <- herald:::.op_meta()

# Format the arg_schema for each op into a compact one-line summary so
# rule authors can see required arg names at a glance. We pull from
# .OP_META directly because the tibble form drops the nested arg list.
.format_args <- function(name) {
  meta <- herald:::.op_meta(name)
  schema <- meta$arg_schema
  if (length(schema) == 0L) {
    return("")
  }
  parts <- vapply(names(schema), function(arg) {
    spec <- schema[[arg]]
    type <- spec$type %||% "any"
    required <- isTRUE(spec$required)
    if (required) {
      sprintf("%s:%s*", arg, type)
    } else {
      sprintf("%s:%s", arg, type)
    }
  }, character(1))
  paste(parts, collapse = ", ")
}

ops$args <- vapply(ops$name, .format_args, character(1))

# Replace NA kind with "(unset)" so grouping does not drop those rows.
ops$kind <- ifelse(is.na(ops$kind) | !nzchar(ops$kind), "(unset)", ops$kind)
```

Total registered operators: **`r nrow(ops)`**.

```{r render-catalog, results = "asis"}
kind_order <- c(
  "set", "compare", "existence", "temporal", "cross", "string", "spec",
  "(unset)"
)
present_kinds <- intersect(kind_order, sort(unique(ops$kind)))
extra_kinds <- setdiff(sort(unique(ops$kind)), present_kinds)
all_kinds <- c(present_kinds, extra_kinds)

for (k in all_kinds) {
  rows <- ops[ops$kind == k, , drop = FALSE]
  if (nrow(rows) == 0L) next
  cat(sprintf("\n\n## %s (%d)\n\n", k, nrow(rows)))
  tbl <- data.frame(
    name = rows$name,
    summary = rows$summary,
    args = rows$args,
    cost = rows$cost_hint,
    `NA ok` = ifelse(rows$returns_na_ok, "yes", "no"),
    check.names = FALSE,
    stringsAsFactors = FALSE
  )
  print(knitr::kable(tbl, format = "html", row.names = FALSE))
}
```

## Reading the table

- **name** -- the registry key referenced by rule YAML under the `op:`
  field.
- **summary** -- one-line description from the op's registration block.
- **args** -- argument names with type tags. A trailing `*` marks a
  required argument; argless ops show an empty cell.
- **cost** -- big-O hint for the inner loop. `O(n)` ops scale linearly
  with dataset rows; `O(n*m)` ops scan a reference dataset.
- **NA ok** -- whether the op emits `NA` (advisory) under indeterminate
  conditions such as a missing required argument or a partially typed
  column.

## See also

- [Architecture](../architecture.html) -- where operators sit in the
  layer stack.
- [Rule coverage](../rule-coverage.html) -- how compiled rules are
  scoped to ops and standards.
- `vignette("validation-reporting", package = "herald")` -- end-to-end
  validation flow.
